---
title: "Alabama Tornado Analysis"
author: "Owen Widdis"
date: "2025-12-01"
output: github_document
---
# Background

In 2008, while on a family trip to the Wisconsin Dells, we were forced to take shelter in a roadside ditch to escape an oncoming tornado. Thankfully, we emerged unharmed, but the experience left a lasting impression on me. As I grew older, that encounter sparked an interest in tornadoes and their effects on infrastructure, the environment, and the lives of those in their path. 

# Research Question

With growing concern about climate change and its potential impact on storm intensity, I became interested in exploring how shifts in atmospheric conditions have influenced tornadoes. Specifically, I wanted to investigate whether or not there is an affect on inflation-adjusted damage costs from tornadoes in Alabama from 1950 - 2025.

### Intensity Variables

To measure the intensity of a given storm, I considered several variables. In 1971, Ted Fujita developed the Fujita Scale to distinguish between weak and strong tornadoes. This scale was quickly adopted by the meteorological community; however, long-term use revealed several inconsistencies and inaccuracies in storm ratings. Researchers at Texas Tech University proposed the Enhanced Fujita (EF) Scale in 2004, building upon the original framework, which rates storms on a scale of 0-5, with 5 being the most intensive. The EF Scale uses damage indicators and degrees of damage to estimate wind speeds, which are then used to assign a storm’s rating. The National Weather Service officially adopted the EF Scale in 2007.

It is important to note that the dataset used for this project (the Storm Events Database from the National Oceanic and Atmospheric Administration) includes both F and EF Scale ratings (F Scale: 1950–February 1, 2007; EF Scale: February 1, 2007–2025). Tornadoes that occurred prior to the development of the Fujita Scale in 1971 were rated retroactively using newspaper reports, eyewitness accounts, and other available documentation describing storm impacts.

Second, I examined damage estimates from each storm across two variables: property damage and crop damage. To account for inflation, I used the quantmod library to access Federal Reserve Consumer Price Index (CPI) data for each month from 1950 to 2025 and adjusted all damage estimates accordingly.

### Focus Area

I selected Alabama as my focus region for several key reasons. First and foremost, Alabama experiences a high frequency of tornadic activity. Second, the state has been affected by numerous high-impact and costly tornado events, largely due to its relatively dense human and built infrastructure compared to other tornado-prone regions such as Kansas and Oklahoma, which are more predominantly rural and agricultural. As a result, Alabama represents a compelling case study, as it encompasses a mix of agricultural land use and concentrated population centers, allowing for an examination of tornado impacts across varied infrastructure types.

# Methods

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Kendall)
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
library(quantmod)
library(scales)
library(ggplot2)
library(viridis)  # for nice color scales
library(sf)

options(scipen = 999)

```

### Load data 

```{r}

# Load CSV
tornado_df <- read_csv("C:/Users/walke/Documents/umich/courses/SI544/final_proj/tornado_analysis/merged/mega_tornado_filtered.csv")

# Convert numeric columns
numeric_cols <- c(
  "TOR_LENGTH", "TOR_WIDTH",
  "DEATHS_DIRECT", "INJURIES_DIRECT",
  "DAMAGE_PROPERTY_NUM", "DAMAGE_CROPS_NUM"
)

tornado_df <- tornado_df %>%
  mutate(across(all_of(numeric_cols), ~as.numeric(.)))

# Convert dates + create DATE column for joining
tornado_df <- tornado_df %>%
  mutate(
    BEGIN_DATE = mdy(BEGIN_DATE),
    END_DATE   = mdy(END_DATE),
    DATE       = as.Date(BEGIN_DATE),
    YEAR       = year(BEGIN_DATE)
  )

# rm columns
tornado_df <- tornado_df %>%
  select(
    -EVENT_ID,
    -CZ_NAME_STR,
    -EVENT_TYPE
  )

```
### Load CPI data and join it with Tornado dataframe, then cleanup dataframe.

```{r}

# Download CPI from FRED (monthly)
getSymbols("CPIAUCSL", src = "FRED")

# Convert to tibble
cpi_df <- CPIAUCSL %>%
  data.frame(date = index(.)) %>%
  as_tibble() %>%
  rename(CPIAUCSL = CPIAUCSL)

# 2025 CPI target
target_cpi <- 324.368

# Inflation adjustment factor
cpi_df <- cpi_df %>%
  mutate(adj_factor = target_cpi / CPIAUCSL)

tornado_df <- tornado_df %>%
  mutate(
    DATE = as.Date(BEGIN_DATE),
    DATE_MONTH = floor_date(DATE, unit = "month")
  )

# Left join by DATE (tornado) and date (CPI)
tornado_df <- tornado_df %>%
  left_join(cpi_df, by = c("DATE_MONTH" = "date"))

tornado_df <- tornado_df %>%
  mutate(
    DAMAGE_PROPERTY_2025 = DAMAGE_PROPERTY_NUM * adj_factor,
    DAMAGE_CROPS_2025    = DAMAGE_CROPS_NUM * adj_factor,
    DAMAGE_PROPERTY_2025_DOL = dollar(DAMAGE_PROPERTY_2025),
    DAMAGE_CROPS_2025_DOL    = dollar(DAMAGE_CROPS_2025)
  )

# Clean tornado dataset with uppercase column names
tornado_clean <- tornado_df %>%
  # Keep only relevant columns (uppercase)
  select(
    DATE, YEAR,
    BEGIN_DATE, END_DATE,
    BEGIN_LAT, BEGIN_LON,
    END_LAT, END_LON, TOR_F_SCALE,
    TOR_LENGTH, TOR_WIDTH,
    DEATHS_DIRECT, INJURIES_DIRECT,
    DAMAGE_PROPERTY_NUM, DAMAGE_CROPS_NUM,
    DAMAGE_PROPERTY_2025, DAMAGE_CROPS_2025
  ) %>%
  # Rename columns to uppercase + more readable
  rename(
    START_DATE = BEGIN_DATE,
    END_DATE   = END_DATE,
    START_LAT  = BEGIN_LAT,
    START_LON  = BEGIN_LON,
    END_LAT    = END_LAT,
    END_LON    = END_LON,
    TORNADO_LENGTH = TOR_LENGTH,
    TORNADO_WIDTH  = TOR_WIDTH,
    DEATHS  = DEATHS_DIRECT,
    INJURIES = INJURIES_DIRECT,
    DAMAGE_PROPERTY = DAMAGE_PROPERTY_NUM,
    DAMAGE_CROPS    = DAMAGE_CROPS_NUM,
    DAMAGE_PROPERTY_2025_USD = DAMAGE_PROPERTY_2025,
    DAMAGE_CROPS_2025_USD    = DAMAGE_CROPS_2025
  ) %>%
  # Add formatted dollar versions for display
  mutate(
    DAMAGE_PROPERTY_2025_DOL = dollar(DAMAGE_PROPERTY_2025_USD),
    DAMAGE_CROPS_2025_DOL    = dollar(DAMAGE_CROPS_2025_USD)
  ) %>%
  # Sort by DATE ascending
  arrange(DATE)
tornado_clean <- tornado_clean %>%
  mutate(
    TOR_F_SCALE_NUM = as.numeric(str_extract(TOR_F_SCALE, "\\d+"))
  )

# Preview
tail(tornado_clean, 1000)

```

### Exploratory Data Analysis

The following visualizations give a glimpse into the summary statistics of the data

#### Annual Damage Data Frame

Firstly, I created a data frame that displays the intensity variables of a given year, showing the year, total damages, amount of *recorded* tornadoes, and the mean intensity rating. 

```{r}

annual_damage <- tornado_clean %>%
  group_by(YEAR) %>%
  summarise(
    total_property_damage = sum(DAMAGE_PROPERTY_2025_USD, na.rm = TRUE),
    total_crop_damage     = sum(DAMAGE_CROPS_2025_USD, na.rm = TRUE),
    full_damage_cost      = sum(total_property_damage + total_crop_damage, na.rm = TRUE),
    tornado_count         = n(),
    mean_intensity        = mean(TOR_F_SCALE_NUM, na.rm = TRUE)
  )

annual_damage

```

#### Time trend of Damage Costs
Second, this visualization presents annual total tornado damage costs. While most years are characterized by relatively modest losses, several extreme outliers are evident, most notably 2011, which stands out as an anomalously destructive year due to widespread and unusually intense tornado activity in Alabama and across the United States.

```{r}

ggplot(annual_damage, aes(x = YEAR, y = full_damage_cost)) +
  geom_line() +
  scale_y_continuous(labels = dollar) +
  coord_cartesian(ylim = c(0, 2000000000)) +
  labs(
    title = "Annual Inflation-Adjusted Damage Costs from Tornadoes in Alabama",
    x = "Year",
    y = "Total Damage (2025 USD)"
  ) +
  theme_minimal()

```

#### Damage Cost Distributions
The following visualizations show the distribution of damage cost ranges for each year. The distibution is highly right skewed, meaning there are only a few years that have experienced catastophic damage amounts, while the majority of years have relatively low damage amounts. To better visualize the distribution and support regression analysis, the second visualization shows the damage values log-transformed, which reduces the influence of extreme outliers and allows for clearer comparison across typical years. The log-scaled x-axis represents orders of magnitude of annual damage, allowing years with vastly different loss totals to be compared more clearly.

```{r}

ggplot(annual_damage, aes(x = full_damage_cost)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(labels = dollar) +
  labs(
    title = "Distribution of Annual Tornado Damage",
    x = "Annual Property Damage (2025 USD)",
    y = "Count"
  ) +
  theme_minimal()

ggplot(annual_damage, aes(x = log(full_damage_cost + 1))) +
  geom_histogram(bins = 40) +
  labs(
    title = "Log-Transformed Distribution of Annual Property Damage",
    x = "Log(Property Damage + 1)",
    y = "Count"
  ) +
  theme_minimal()

```

#### Boxplot
The below visual shows a boxplot that shows how the rating of tornadoes relates to the amount of damage cost the storm produced. The plot shows that median damage generally increases with tornado intensity, but there is substantial overlap between categories. Lower-rated tornadoes occasionally produce high damage, often due to factors such as population density and infrastructure exposure, while not all high-intensity tornadoes result in extreme damage. This highlights that while tornado intensity is strongly related to damage potential, exposure and vulnerability play a critical role in determining realized economic impact.

```{r}

ggplot(tornado_clean %>% filter(DAMAGE_PROPERTY_2025_USD > 0 | DAMAGE_CROPS_2025_USD > 0),
       aes(x = factor(TOR_F_SCALE_NUM), y = DAMAGE_PROPERTY_2025_USD + DAMAGE_CROPS_2025_USD)) +
  geom_boxplot() +
  scale_y_log10(labels = dollar) +
  labs(
    title = "Tornado Damage by F/EF Rating",
    x = "F / EF Rating",
    y = "Property Damage (2025 USD, log scale)"
  ) +
  theme_minimal()

```

Lastly, this dataframe shows overall summary statistics.

```{r}

annual_damage %>%
  summarise(
    mean_annual_damage   = dollar(mean(full_damage_cost)),
    median_annual_damage = dollar(median(full_damage_cost)),
    max_annual_damage    = dollar(max(full_damage_cost)),
    mean_tornadoes_per_year = round(mean(tornado_count), 1)
  )

```

### Regression

To understand the relationship between tornado characteristics and economic impacts, I used a linear regression model to predict the logarithm of annual inflation-adjusted tornado damage in Alabama from 1950 to 2025. Independent variables included year, mean tornado intensity, and the total number of tornadoes per year. The model was statistically significant overall (p < 0.001), explaining approximately 17% of the variation in annual damage. The results show that tornado frequency is indicative of damage increase, with each additional tornado adds an estimated 2.8% increase in annual costs. Mean tornado intensity exhibited a positive but statistically insignificant relationship with damage, suggesting that the number of tornadoes may play a larger role in determining annual economic impacts than average storm strength alone.

```{r}

annual_damage <- annual_damage %>%
  mutate(
    log_full_damage = log(full_damage_cost + 1)
  )

model1 <- lm(
  log_full_damage ~ YEAR + mean_intensity + tornado_count,
  data = annual_damage
)

summary(model1)

```


#### Regression Visualizations

The following visualizations are based on the regression model. The first visual shows the observed vs predicted values predicted by the model. The axes are both logarithmically scaled to account for outlying years. Points closer to the diagonal line indicate years where the model predicts damage well, while large deviations reflect years dominated by extreme or localized events.

The second visualization shows the relationship between the number of tornadoes occurring in a given year and the log adjusted damage costs. Each point represents a single year, while the fitted regression line and confidence interval depict the estimated marginal effect of tornado frequency on economic losses. The positive slope indicates that years that had more tornadoes are associated with greater total damage, supporting the results that event frequency is a primary driver of annual monetary impacts.

```{r}

annual_damage <- annual_damage %>%
  mutate(
    predicted_log_damage = predict(
      lm(log_full_damage ~ YEAR + mean_intensity + tornado_count,
         data = annual_damage)
    ),
    predicted_damage = exp(predicted_log_damage)
  )

ggplot(annual_damage, aes(x = full_damage_cost, y = predicted_damage)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = dollar) +
  scale_y_log10(labels = dollar) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Observed vs. Predicted Annual Tornado Damage",
    x = "Observed Damage (2025 USD, log scale)",
    y = "Predicted Damage (2025 USD, log scale)"
  ) +
  theme_minimal()

ggplot(annual_damage, aes(x = tornado_count, y = log_full_damage)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Relationship Between Tornado Frequency and Log Annual Damage",
    x = "Number of Tornadoes per Year",
    y = "Log Annual Damage (2025 USD)"
  ) +
  theme_minimal()


```

### Mann-Kendall Test

After doing research on other climate and hydrology research projects, I believed it would be beneficial to run and Mann-Kendall test to determine whether or not there is a monotonic trend in damage costs from tornado damages over time. Different than linear regression analyses, the MK test does not assume linearity and is not prone to outliers or extreme years, as it is just measuring if there is an increase or decrease in a given variable. 

The below visual shows annual inflation-adjusted tornado damage in Alabama from 1950 to 2025 on a log adjusted scale, along with a locally weighted smoothing curve to highlight long-term patterns. While year-to-year damage varies substantially and is dominated by a small number of extreme years, the smoothed trend suggests a modest decline in damage over time. This visual pattern aligns with results from the Mann–Kendall test, which detects a weak but statistically significant decreasing monotonic trend (tau = –0.175, 2 sided pvalue = 0.026).

```{r}
mk_test <- MannKendall(annual_damage$log_full_damage)

ggplot(annual_damage, aes(x = YEAR, y = log_full_damage)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(
    title = "Long-Term Trend in Log-Adjusted Annual Tornado Damage",
    x = "Year",
    y = "Log Annual Damage (2025 USD)"
  ) +
  theme_minimal()

mk_test

```

#### Interpretation of MK Test

The Tau value measures the strength and direction of a trend by comparing the relative ordering of values over time. The negative value of tau tells us that a decrease in annual damage slightly outnumbers increases across the study period, while the small magnitude reflects a weak trend rather than a steep decline. Alongside the linear regression results, it suggests that inflation-adjusted tornado damages have not increased over time and may have ever so slightly declined, even as interannual variability remains high.

These results suggest that long-term economic impacts from tornadoes in Alabama are not characterized by increasing inflation-adjusted damages over time. While climate change may influence the atmospheric conditions associated with severe storms, the observed damage patterns likely due to the combined effects of improved forecasting, warning systems, building practices, and emergency response, which may have offset potential increases in hazard-related risk. Additionally, the data being used spans across ~75 years where many surveying techniques were used and adapted during this period, making it succeptible to inconsistencies and data that is not definitively reflective of the true population.

### Geospatial Visualization

NOAA storm event data provides the starting and ending latitude and longitude values for a given tornado. This piqued my interest as it allows for a great opportunity to create a spatial visualization of tornadic impact on Alabama.

I created a shapefile and exported it as a GEOJSON file to be able to work with the data on ArcGIS online. 

```{r}
tornado_clean_filtered <- tornado_clean %>%
  filter(!is.na(START_LAT) & !is.na(START_LON) &
         !is.na(END_LAT) & !is.na(END_LON))

# Create sf LINEString
tornado_paths <- tornado_clean_filtered %>%
  rowwise() %>%
  mutate(
    geometry = list(
      st_linestring(
        matrix(
          c(START_LON, START_LAT,
            END_LON,   END_LAT),
          ncol = 2,
          byrow = TRUE
        )
      )
    )
  ) %>%
  st_as_sf(crs = 4326)

print(tornado_paths)


st_write(tornado_paths, "tornado_paths_alabama.geojson", driver = "GeoJSON")




```

### Concluding Thoughts

This project applied data-driven methods to evaluate long-term trends in inflation-adjusted tornado damage in Alabama from 1950 to 2025. By integrating historical tornado records with Consumer Price Index data and aggregating individual events to annual summaries, I hoped to determine whether economic impacts from tornadoes have increased over time. 

The summary statistics and visualizations showed that tornado damage is extremely skewed by a few years, with a small number of years accounting for a majority of the proportion of total losses. These skews motivated me to  use log transformations to deliver a more comparable analysis across the study period. This choice highlights the importance of using adjusted scales that are often necessary when working with hazard and disaster loss data.

The regression model results indicate that annual tornado damage is more associated with event frequency. After controlling for year, mean tornado intensity, and the number of tornadoes per year, only tornado frequency exhibited a statistically significant relationship with annual damage. The absence of a significant positive time trend suggests that inflation-adjusted tornado damages in Alabama have not increased  over the study period. These findings were reinforced by the Mann–Kendall trend test, which detected a weak but statistically significant decreasing monotonic trend in log-adjusted damage. Together, these results suggest that temporal trends in observed economic impacts are not dominated by increasing hazard intensity alone.

These findings highlight the distinction between hazard and impact. Economic damage is not solely a function of storm strength, but also of where tornadoes occur relative to population centers, infrastructure density, land use, and built environment vulnerability. Over the study period, there have been significant improvements in forecasting, warning systems, and infrastructure that have likely contributed to the dampening long-term damage trends, even as atmospheric conditions evolve from climate change.

#### Future Directions

This study opens several directions for future research. One  extension would be to expand the scope beyond a single state and examine regional or national patterns in tornado damage. A broader study region would allow for comparative analysis across regions with differing land use, population density, and climate trends that would enable more robust inference about spatial variability in tornado impacts.

Additionally, incorporating geospatial variables such as population density, urban–rural classification, land cover, or proximity to critical infrastructure could improve results. Event-level spatial modeling, rather than annual aggregation, would also allow for finer-scale analysis of how tornado paths intersect with vulnerable areas.

Overall, this project demonstrates how combining statistical modeling and temporal trend analysis can yield insights into long-term hazard impacts. 
